{\bf {\Large Improving estimates of ecosystem metabolism by reducing effects of tidal advection on dissolved oxygen time series}}

### 2015-06-04 intro.Rnw
incorporated Jim's comments on revision, need to proof

Ecosystem metabolism describes the balance between production and respiration processes that create and consume organic matter \citep{Kemp12,Needoba12}.  Light exposure experiments of water samples collected at discrete locations and times have traditionally been used to measure metabolic activity \citep{Gaarder27}.  Although highly controlled and precise, bottle-based methods are labor-intensive and not scalable to describe entire ecosystem rates.  Bottle-based methods may also only reliably estimate production and respiration associated with planktonic processes, whereas significant contributions to ecosystem metabolism can arise from other habitats, such as the benthos or intertidal marshes.  By contrast, open-water techniques have been increasingly used to estimate whole system metabolism given the availability of long-term, continuous time series of dissolved oxygen \citep{Odum56,Davanzo96}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration attributed to all ecosystem components. Open-water estimates also provide a more accessible means of tracking ecosystem change over time as compared to discrete sampling events with bottle-based approaches. Although metabolic rates vary naturally at different spatial and temporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production (\citealt{Nixon95}; \citetalias{NRC00}).  Reliable estimates of whole ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem function.

@@

This article describes the development and application of a method for improving estimates of ecosytem metabolism computed from \ac{DO} time series by reducing the effect of tidal advection on continuous \ac{DO} observations.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  The model is based on the recognition that daily fluctuations in \ac{DO} caused by metabolism are associated with the solar cycle, whereas other fluctuations in estuaries are likely associated with cyclical water level changes that generally exhibit pregression relative to the solar cycle.  The weighted regression model was applied, rather than methods commonly used for detiding in physical oceanography, to allow for the complex and dynamic patterns of \ac{DO} changes relative to advection.  The method targets the periodicity of the tidal component as an explicit variable during model fitting, which allows the ability to isolate the biological component of the dissolved oxygen time series.  As a result, the weighted regression approach can preserve the true biological signal in the output rather than risking removal of both the biological and physical components as with traditional deconvulution methods.  This approach offers a distinct advantage for estuaries where the magnitude of tidal effects on water quality observations can be severe.  During initial method development, we used simulated \ac{DO} time series with known characteristics to evaluate the ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Subsequently, the simulation results informed further development of the method as applied to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  In all examples, tidal height is used as a proxy for advection because it was recorded with the \ac{NERRS} data.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous stations along a tidal axis), the model determines the advective effect empirically.

>>>>

Ecosystem metabolism describes the balance between production and respiration processes that create and consume organic matter \citep{Kemp12,Needoba12}.  Light exposure experiments of water samples collected at discrete locations and times have traditionally been used to measure metabolic activity \citep{Gaarder27}.  Although highly controlled and precise, bottle-based methods are labor-intensive and not scalable to describe entire ecosystem rates.  Bottle-based methods may also only reliably estimate production and respiration associated with planktonic processes, whereas significant contributions to ecosystem metabolism can arise from other habitats, such as the benthos or intertidal marshes.  By contrast, open-water techniques have been increasingly used to estimate whole system metabolism given the availability of long-term, continuous time series of dissolved oxygen \citep{Odum56,Davanzo96}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration attributed to all ecosystem components. Open-water estimates also provide a practical means of capturing events and ecosystem changes over time as compared to discrete sampling events with bottle-based approaches because monitors are present continuously.  While bottle-based measurements offer advantages in precision, it is not possible to sustain bottle measurements day after day, making it unlikely that sampling will occur during an unanticipated short-term event.  Moreover, more variability is captured in continuous data and statistical power for resolving long-term changes is greater compared to periodic measurements.  Although metabolic rates vary naturally at different spatial and temporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production (\citealt{Nixon95}; \citetalias{NRC00}).  Reliable estimates of whole ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem function.

@@

This article describes the development and application of a method for improving estimates of ecosytem metabolism computed from \ac{DO} time series by reducing the effect of tidal advection on continuous \ac{DO} observations.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  The model is based on the recognition that daily fluctuations in \ac{DO} caused by metabolism are associated with the solar cycle, whereas other fluctuations in estuaries are likely associated with cyclical water level changes that generally exhibit pregression relative to the solar cycle.  The weighted regression model was applied, rather than methods commonly used for detiding in physical oceanography, to allow for the complex and dynamic patterns of \ac{DO} changes relative to advection.  The method targets the tidal component as an explicit variable using dynamic model parameters that change through time.  This approach makes it possible to estimate and reduce the effect of advection on the DO time series, even if this effect varies through time due to changes in metabolic rates.  Local fitting provides an advantage relative to traditional deconvulution methods. The method is most useful in estuaries where tidal effects interact with horizontal differences in metabolism and associated \ac{DO} concentrations.  Although the method could be applied in open ocean environments, it may be less useful because horizontal \ac{DO} gradients may be less in such systems, such that it could be sufficient to simply neglect advection as has commonly been done in the past. During initial method development, we used simulated \ac{DO} time series with known characteristics to evaluate the ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Subsequently, the simulation results informed further development of the method as applied to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  In all examples, tidal height is used as a proxy for advection because it was recorded with the \ac{NERRS} data.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous stations along a tidal axis), the model determines the advective effect empirically.

### 2015-06-04 mat_proc.Rnw
incorporated Jim's comments on revision, need to proof

Tidal height is used as a proxy for advection because the model is empirical and only requires a variable that is correlated with the true measure of interest.  An important distinction between tidal height and advection is that each variable could provide different information about tidal regime.  For example, tidal heights at the minimum and maximum of the range for standing wave tides may be associated with periods of low advection when water masses are not moving rapidly past the sensor.  The relationship between tidal height and advection across multiple environments may be complex.  The weighted regression model is suited for application to different locations because it needs only a variable that indicates a particular point in the tidal cycle.  The specific influence of advection on the \ac{DO} time series is isolated using the moving window approach such that the relationship between tidal height and advection does not need to be known.

>>>>

Tidal height is used as a proxy for advection because this measurement is widely available, easy to measure, and meets the model requirements, which is a variable that is correlated locally in time with \ac{DO} variations attributable to advection.  Importantly, tide height need not predict the actual magnitude of advection.  This is important because, across estuarine locations, the relationship between tide height and tidal currents can be variable in both magnitude and phase, depending on local characteristics.  The weighted regression model is well-suited for the situation in estuaries because it empirically fits the local relationship between tide height and \ac{DO}, such that no a priori knowledge of the relationship between tide height and either the magnitude of advection or the horizontal gradient in \ac{DO} is needed.

### 2015-06-04 swmp_noise.Rnw
incorporated Jim's comments on revision, need to proof

% example of creating simulated time series
\centering\vspace*{\fill}
<<do_sim, eval = T, echo = F, cache = T, fig.width = 7, fig.height = 6, eval.after='fig.cap', fig.cap = 'Example of each component of a simulated \\ac{DO} time series for testing weighted regression.  The time series were created using \\cref{do_obs,do_bio,do_unc,do_obs_all,do_sin,do_unc_n,deltdo,deltx,do_advp,do_adv}. Yellow indicates a twelve hour daylight period beginning at 630 each day.'>>=
# create time vector
vec <- c('2014-05-01 00:00:00', '2014-05-31 00:00:00')
vec <- as.POSIXct(vec, format = '%Y-%m-%d %H:%M:%S')
vec <- seq(vec[1], vec[2], by = 60*30)

@@

% example of error surfaces 
\centering\vspace*{\fill}
<<err_surf1, fig.height = 3.5, fig.width = 5.5, cache = T, echo = F, eval = T, fig.cap = "Heat maps of correlations and errors (\\ac{RMSE}) for filtered \\ac{DO} time series ($DO_{dtd}$) from weighted regression with `true' biological \\ac{DO} ($DO_{bio}$) for varying simulation parameters: strength of tidal association with \\ac{DO} signal ($DO_{adv}$), amount of process uncertainty ($\\varepsilon_{pro}$), amount of observation uncertainty ($\\varepsilon_{obs}$), and strength of diel \\ac{DO} component ($DO_{die}$).  Each tile represents the correlation or error from results for a given combination of simulation parameters averaged for all window widths (as in \\cref{fig:err_surf2}).  See \\cref{tab:dtd_perf1} for a summary of combined results for each unique parameter.">>=

@@

% example of error surfaces 
\centering\vspace*{\fill}
<<err_surf2, fig.height = 2.33, fig.width = 6, cache = T, echo = F, eval = T, fig.cap = "Heat maps of correlations and errors (\\ac{RMSE}) for filtered \\ac{DO} time series ($DO_{dtd}$) from weighted regression with `true' biological \\ac{DO} ($DO_{bio}$) for varying half window widths: days, hour of day, and proportion of tidal range.  Each tile represents the correlation or error from results for a given combination of window widths averaged for all simulation parameters (as in \\cref{fig:err_surf1}).  See \\cref{tab:dtd_perf2} for a summary of combined results for each unique window.">>=

@@

% summary of simulation performance for detided and biological, sim parameters
<<dtd_perf1, cache = F, echo = F, eval = T, results = 'asis'>>=

# uses 'mod_perf' from, see readme
load('data/mod_perf.RData')

# calculate mean, se for dtd_cor and dtd_err
dtd_perf <- llply(names(mod_perf)[2:8], # note indexing
  .fun = function(x) {
    ddply(
      mod_perf,
      .variables = x,
      .fun = function(y) {
        c(
          min(y$cor),
          quantile(y$cor, 0.25),
          median(y$cor),
          quantile(y$cor, 0.75),
          max(y$cor),
          min(y$err),
          quantile(y$err, 0.25),
          median(y$err),
          quantile(y$err, 0.75),
          max(y$err)
        )
    })}
  )

# rename first column for combining, create column for parameter,  then combine
dtd_perf <- lapply(dtd_perf, 
  function(x) { 
    Parameter <- names(x)[1]
    names(x)[1] <- 'Level'
    x[,1] <- factor(x[,1])
    Parameter <- rep(Parameter, nrow(x))
    x <- data.frame(Parameter, x)
    x
    }
  )
dtd_perf <- do.call('rbind', dtd_perf)
names(dtd_perf) <- c('Parameter', 'Level', 'Min cor', 'Low cor', 'Median cor', 'High cor', 'Max cor', 'Min err', 'Low err', 'Median err', 'High err', 'Max err')

# subset by DO time series conditions
to_tab <-dtd_perf[!dtd_perf$Parameter %in% c('dec_time', 'hour', 'Tide'), ]

# names for row categories in table
Parms <- c('$\\boldsymbol{DO_{die}}$', 
  '$\\boldsymbol{DO_{adv}}$',  
  '$\\boldsymbol{\\varepsilon_{pro}}$', 
  '$\\boldsymbol{\\varepsilon_{obs}}$'
  )
tab <- form_fun(to_tab[, -c(1:2)])
rows <- to_tab[, 2]
cap.val<-'Summary (range, median, quartiles) of correlations and error estimates comparing filtered and biological \\ac{DO} time series for different simulation parameters ($DO_{die}$, $DO_{adv}$, $\\varepsilon_{pro}$, $\\varepsilon_{obs}$).  Values represent averages from multiple simulations with common parameters.  For example, row one is a summary of all simulations for which the diel \\ac{DO} component was zero ($n=729$).  See \\cref{fig:err_surf1} for results of all parameter combinations.'

# foot.val<-'\\footnotesize \\textit{Note:} test'

latex( 
  tab,
  file = '',
  where = 'h',
  rowlabel = 'Parameter',
#   insert.bottom = foot.val,
  caption = cap.val,
  caption.loc = 'top',
  rgroup = Parms,
  n.rgroup = rep(3, 4),
  cgroup = c('Correlation', 'RMSE'),
  n.cgroup = c(5, 5),
  rowname = rows,
  colheads=rep(c('Min', '25\\textsuperscript{th}', 'Median', '75\\textsuperscript{th}', 'Max'),2),
  label = 'tab:dtd_perf1'
  )
@

% summary of simulation performance for detided and biological, window widths
<<dtd_perf2, cache = F, echo = F, eval = T, results = 'asis'>>=

# uses 'mod_perf' from 'dep_data.rnw'
load('data/mod_perf.RData')

# calculate median, se for dtd_cor and dtd_err
dtd_perf <- llply(names(mod_perf)[1:8],
  .fun = function(x) {
    ddply(
      mod_perf,
      .variables = x,
      .fun = function(y) {
        c(
          min(y$cor),
          quantile(y$cor, 0.25),
          median(y$cor),
          quantile(y$cor, 0.75),
          max(y$cor),
          min(y$err),
          quantile(y$err, 0.25),
          median(y$err),
          quantile(y$err, 0.75),
          max(y$err)
        )
    })}
  )

# rename first column for combining, create column for parameter,  then combine
dtd_perf <- lapply(dtd_perf, 
  function(x) { 
    Parameter <- names(x)[1]
    names(x)[1] <- 'Level'
    x[,1] <- factor(x[,1])
    Parameter <- rep(Parameter, nrow(x))
    x <- data.frame(Parameter, x)
    x
    }
  )
dtd_perf <- do.call('rbind', dtd_perf)
names(dtd_perf) <- c('Parameter', 'Level', 'Min cor', 'Low cor', 'Median cor', 'High cor', 'Max cor', 'Min err', 'Low err', 'Median err', 'High err', 'Max err')

# subset by windows
to_tab <-dtd_perf[dtd_perf$Parameter %in% c('dec_time', 'hour', 'Tide'), ]

# names for row categories in table
Parms <- c('Days', 'Hours', 'Tide')

tab <- form_fun(to_tab[, -c(1:2)])
rows <- to_tab[, 2]
cap.val<-'Summary (range, median, quartiles) of correlations and error estimates comparing filtered and biological \\ac{DO} time series for simulations using different half window widths in the weighted regressions (days, hours, and proportion of tidal range).  Values represent averages from multiple simulations with common window values. For example, row one is a summary of all simulations for which the half window width was one day ($n=729$). See \\cref{fig:err_surf2} for results of all window combinations.'

# foot.val<-'\\footnotesize \\textit{Note:} test'

latex( 
  tab,
  file = '',
  rowlabel = 'Window',
#   insert.bottom = foot.val,
  caption = cap.val,
  caption.loc = 'top',
  rgroup = Parms,
  n.rgroup = rep(3, 3),
  cgroup = c('Correlation', 'RMSE'),
  n.cgroup = c(5, 5),
  rowname = rows,
  colheads=rep(c('Min', '25\\textsuperscript{th}',  'Median', '75\\textsuperscript{th}', 'Max'),2),
  label = 'tab:dtd_perf2'
  )
@

@@

foot.val<-'\\footnotesize\\textsuperscript{\\textit{a}}Pg: gross production, Rt: respiration, NEM: net ecosystem metabolism, estimated using methods described herein with observed data'

latex( 
  tab,
  file = '',
  rowlabel = 'Site',
  insert.bottom = foot.val,
  caption = cap.val,
  caption.loc = 'top',
  cgroup = c('Tidal amplitude', 'Water quality', 'Metabolism\\textsuperscript{\\textit{a}}'),
  n.cgroup = c(4, 4, 3),
  rowname = rows,
  colheads=c('O1', 'P1', 'M2', 'S2', 'DO', 'Chl', 'Sal', 'Temp', 'Pg', 'Rt', 'NEM'),
  label = 'tab:case_att'
  )

@@

@@

foot.val<-'\\textsuperscript{\\textit{a}}Pg: gross production, Rt: respiration' 
latex( 
  tab,
  file = '',
  rowlabel = 'Site',
  insert.bottom = foot.val,
  caption = cap.val,
  caption.loc = 'top',
  rgroup = unique(to_tab$Site),
  n.rgroup = rep(2, 4),
  cgroup = c('Pg\\textsuperscript{\\textit{a}}', 'Rt'),
  n.cgroup = c(3, 3),
  rowname = rows,
  colheads=c(rep(c('Mean','SD', 'Anom'),2), c('Mean', 'SD')),
  label = 'tab:case_res'
  )
@

>>>>

% example of creating simulated time series
\centering\vspace*{\fill}
<<do_sim, eval = T, echo = F, cache = T, fig.width = 7, fig.height = 6, eval.after='fig.cap', fig.cap = 'Example of each component of a simulated \\ac{DO} time series for testing weighted regression.  The time series were created using \\cref{do_obs,do_bio,do_unc,do_obs_all,do_sin,do_unc_n,deltdo,do_advp}. Yellow indicates a twelve hour daylight period beginning at 630 each day.'>>=
# create time vector
vec <- c('2014-05-01 00:00:00', '2014-05-31 00:00:00')
vec <- as.POSIXct(vec, format = '%Y-%m-%d %H:%M:%S')
vec <- seq(vec[1], vec[2], by = 60*30)

@@

% example of error surfaces 
\centering\vspace*{\fill}
<<err_surf1, fig.height = 3.5, fig.width = 5.5, cache = T, echo = F, eval = T, fig.cap = "Heat maps of correlations and errors (\\ac{RMSE}) for filtered \\ac{DO} time series ($DO_{dtd}$) from weighted regression with `true' biological \\ac{DO} ($DO_{bio}$) for varying simulation parameters: strength of tidal association with \\ac{DO} signal ($DO_{adv}$), amount of process uncertainty ($\\varepsilon_{pro}$), amount of observation uncertainty ($\\varepsilon_{obs}$), and strength of diel \\ac{DO} component ($DO_{die}$).  Each tile represents the correlation or error from results for a given combination of simulation parameters averaged for all window widths (as in \\cref{fig:err_surf2}).">>=

@@

% example of error surfaces 
\centering\vspace*{\fill}
<<err_surf2, fig.height = 2.33, fig.width = 6, cache = T, echo = F, eval = T, fig.cap = "Heat maps of correlations and errors (\\ac{RMSE}) for filtered \\ac{DO} time series ($DO_{dtd}$) from weighted regression with `true' biological \\ac{DO} ($DO_{bio}$) for varying half window widths: days, hour of day, and proportion of tidal range.  Each tile represents the correlation or error from results for a given combination of window widths averaged for all simulation parameters (as in \\cref{fig:err_surf1}).">>=

@@

@@

foot.val<-'\\footnotesize\\textsuperscript{\\textit{a}}Pg: gross production, Rt: respiration, NEM: net ecosystem metabolism, estimated using methods described herein without removing advection effects.'

latex( 
  tab,
  where = '!h', 
  file = '',
  rowlabel = 'Site',
  insert.bottom = foot.val,
  caption = cap.val,
  caption.loc = 'top',
  cgroup = c('Tidal amplitude', 'Water quality', 'Metabolism\\textsuperscript{\\textit{a}}'),
  n.cgroup = c(4, 4, 3),
  rowname = rows,
  colheads=c('O1', 'P1', 'M2', 'S2', 'DO', 'Chl', 'Sal', 'Temp', 'Pg', 'Rt', 'NEM'),
  label = 'tab:case_att'
  )

@
\clearpage

@@

@
\clearpage

@@

foot.val<-'\\textsuperscript{\\textit{a}}Pg: gross production, Rt: respiration' 
latex( 
  tab,
  file = '',
  rowlabel = 'Site',
  insert.bottom = foot.val,
  caption = cap.val,
  caption.loc = 'top',
  rgroup = unique(to_tab$Site),
  n.rgroup = rep(2, 4),
  cgroup = c('Pg\\textsuperscript{\\textit{a}}', 'Rt'),
  n.cgroup = c(3, 3),
  rowname = rows,
  colheads=c(rep(c('Mean','SD', 'Anom'),2), c('Mean', 'SD')),
  label = 'tab:case_res'
  )
@
\clearpage
